---
title: "Transportability analysis using IOPW with TransportHealthR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transportability analysis using IOPW with TransportHealthR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TransportHealthR)
set.seed(20240501)

expit <- function(x) 1/(1+exp(-x))

# Generate test data for TransportHealth. This data is not meant to be officially put in the package. It is simply to test whether the implemented functions in the package accept the correct inputs and outputs.
generateTestData <- function() {
  # Generate study data
  nStudy <- 1000
  sexStudy <- rbinom(nStudy, 1, 0.5) # Male is 1, so female is baseline
  stressStudy <- rbinom(nStudy, 1, 0.4) # Stressed is 1
  med2Study <- rbinom(nStudy, 1, 0.1) # 1 means taking other med
  percentBodyFatStudy <- rnorm(nStudy, 28 - 13 * sexStudy, 2)
  med1Study <- rbinom(nStudy, 1, expit(0.2 * sexStudy - 0.02 * percentBodyFatStudy + 0.1 * stressStudy))
  sysBloodPressureStudy <- rnorm(nStudy, 100 + 5 * sexStudy + 0.5 * percentBodyFatStudy + 5 * stressStudy -
                                   5 * med1Study + med1Study * (-5 * med2Study + 7 * stressStudy))
  
  # Put all variables together
  studyData <- data.frame(sysBloodPressure = sysBloodPressureStudy, med1 = as.factor(med1Study), sex = as.factor(sexStudy), stress = as.factor(stressStudy), med2 = as.factor(med2Study), percentBodyFat = percentBodyFatStudy)
  
  # Generate target data
  nTarget <- 1500
  sexTarget <- rbinom(nTarget, 1, 0.3) # Male is 1, so female is baseline
  stressTarget <- rbinom(nTarget, 1, 0.7) # Stressed is 1
  med2Target <- rbinom(nTarget, 1, 0.3) # 1 means taking other med
  percentBodyFatTarget <- rnorm(nTarget, 26 - 12 * sexTarget, 2)
  
  # Put all variables together
  targetData <- data.frame(sex = as.factor(sexTarget), stress = as.factor(stressTarget), med2 = as.factor(med2Target), percentBodyFat = percentBodyFatTarget)
  
  return(list(studyData = studyData, targetData = targetData))
}

testData <- generateTestData()
```

# Introduction

Transportability analysis is the statistical problem of estimating causal effects of a treatment on a response for a target population using data gathered from a different population i.e. the study population. It is often used to apply findings from a randomized clinical trial to a population with substantially different characteristics or when it is too difficult to collect data from the population of interest. Marginal causal effects of the study and target populations are often not equal due to the different distributions of effect modifiers between the two populations, so causal effect estimates calculated using samples from the study population are biased for the causal effect of the treatment in the target population. Therefore, effect modification must be accounted for to obtain unbiased causal effect estimates in the target population using data from the study population.

A approach to transportability analysis is inverse odds of participation weighting (IOPW), which weighs observations in the study data by the inverse odds of each observation participating in the study given its effect modifiers in addition to previous weights used for covariate adjustment. The $\mathtt{TransportHealthR}$ package provides functions that streamline this approach.

# Example

Suppose we are interested in estimating the causal effect of a medication on systolic blood pressure in a target population, but we were only able to conduct an observational study using samples from the study population. To obtain unbiased causal effect estimates using the study sample, we account for the following covariates: sex (1 for male), percent body fat, and stress (binary, 1 for stressed). Suppose further that using domain knowledge, we know that the effectiveness of the medication depends on stress levels and whether patients are taking another medication (yes or no), so these variables are effect modifiers. Notice that covariates adjusted for in the study data can also be effect modifiers.